---
title: "Seminar 1"
subtitle: "Human Prion Disease - Statistical Analysis"
author: "CM2018 Group"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
  error: false
fig-width: 11
fig-height: 7
fig-align: center
---

test
2
## Before rendering

Keep `T1_data.csv`, `T3_data.csv` and `T4_data.csv` in the same working directory. Task 2 also requires `CM2018rpackage`. The analyses are connected: Task 2 uses `t1`, while Task 4 uses `t3`.

```{r}
#| echo: false
#| output: false
# ============================================================================
# SEMINAR 1 - TASKS 1, 2, 3 AND 4
# ============================================================================
#
# Keep T1_data.csv, T3_data.csv and T4_data.csv in your working directory.
# In RStudio: Session > Set Working Directory > Choose Directory.
#
# Run the script in sections using Ctrl+Enter.
# Use Source with Echo if you want to run everything and see the outputs.
# Plots appear in the Plots pane: use the back arrow to revisit earlier plots.
# Use Zoom to see figures with several panels.
#
# Only Task 2 requires CM2018rpackage.
#
# Optional: uncomment the following line to save all plots in one PDF.
# Also uncomment dev.off() at the very end.
# pdf("Seminar_1_plots.pdf", width = 12, height = 8)


# ============================================================================
```

# Task 1 - Biomarker investigation

**Research focus:** Do plasma NfL and Tau differ between Healthy and Disease participants?

```{r}
#| echo: false
#| output: false
# TASK 1 - COMPARE HEALTHY AND DISEASE PARTICIPANTS
# ============================================================================

# QUESTION:
# Do NfL and Tau differ between Healthy and Disease participants?
#
# Both biomarkers are continuous and positive.
# Group is categorical, with two levels.
# We assume independent groups because identified matched pairs are not supplied.


```

## 1.1 Import and check the data

Validate the imported dataset before making statistical claims.

```{r}
# 1.1 IMPORT AND CHECK THE DATA

t1 <- read.csv("T1_data.csv")

t1$Group <- factor(
  t1$Group,
  levels = c("Healthy", "Disease")
)

# Healthy is the reference group in lm() and appears first in plots.

str(t1)
head(t1)
summary(t1)

table(t1$Group)
# OUTPUT: Healthy = 25; Disease = 15.

colSums(is.na(t1))
# OUTPUT: no missing values.

anyDuplicated(t1$ID)
# OUTPUT: 0, meaning no repeated participant IDs.

any(t1$NfL <= 0)
any(t1$Tau <= 0)
# OUTPUT: FALSE for both biomarkers.
# All values are positive, so log() can be applied.
#
# Unequal group sizes alone do not establish independence.
# The assignment mentions matching, but its structure is not documented.


```

## 1.2 Descriptive statistics

Describe location and spread on both arithmetic and multiplicative scales.

```{r}
# 1.2 DESCRIPTIVE STATISTICS

aggregate(cbind(NfL, Tau) ~ Group, data = t1, FUN = mean)

# Arithmetic means, in pg/mL:
# Group        NfL       Tau
# Healthy    32.9196    1.6892
# Disease    58.8940    2.9393

aggregate(cbind(NfL, Tau) ~ Group, data = t1, FUN = sd)

# Standard deviations, in pg/mL:
# Group        NfL       Tau
# Healthy    22.7697    1.4210
# Disease    40.4833    2.2963
#
# SD describes variation between participants.
# It is different from the standard error of an estimated mean.

aggregate(cbind(NfL, Tau) ~ Group, data = t1, FUN = median)

# Medians, in pg/mL:
# Group        NfL       Tau
# Healthy     27.02      1.22
# Disease     40.83      2.50

aggregate(
  cbind(NfL, Tau) ~ Group,
  data = t1,
  FUN = function(x) exp(mean(log(x)))
)

# Geometric means, in pg/mL:
# Group        NfL       Tau
# Healthy    27.0902    1.2724
# Disease    46.4691    2.1884
#
# The geometric mean is exp(the mean of the logarithms).
# For values 10, 20 and 40:
# arithmetic mean = 23.33; geometric mean = 20.
#
# The log transformation expresses proportional changes as differences.
# Doubling any value adds log(2).
# Exponentiating a difference between mean logs gives a geometric-mean ratio.
#
# "~ Group" means that the calculations are performed separately by group.
# cbind(NfL, Tau) allows both biomarkers to be summarised together.


```

## 1.3 Compare original and log scales

Compare the original and log scales while retaining every participant.

```{r}
# 1.3 COMPARE ORIGINAL AND LOG SCALES

par(mfrow = c(2, 2))

boxplot(
  NfL ~ Group, data = t1,
  main = "NfL: original scale",
  ylab = "NfL (pg/mL)",
  col = c("lightblue", "salmon"),
  outline = FALSE
)

stripchart(
  NfL ~ Group, data = t1,
  vertical = TRUE, method = "jitter",
  add = TRUE, pch = 16,
  col = rgb(0, 0, 0, 0.5)
)

boxplot(
  log(NfL) ~ Group, data = t1,
  main = "NfL: log scale",
  ylab = "log(NfL)",
  col = c("lightblue", "salmon"),
  outline = FALSE
)

stripchart(
  log(NfL) ~ Group, data = t1,
  vertical = TRUE, method = "jitter",
  add = TRUE, pch = 16,
  col = rgb(0, 0, 0, 0.5)
)

boxplot(
  Tau ~ Group, data = t1,
  main = "Tau: original scale",
  ylab = "Tau (pg/mL)",
  col = c("lightblue", "salmon"),
  outline = FALSE
)

stripchart(
  Tau ~ Group, data = t1,
  vertical = TRUE, method = "jitter",
  add = TRUE, pch = 16,
  col = rgb(0, 0, 0, 0.5)
)

boxplot(
  log(Tau) ~ Group, data = t1,
  main = "Tau: log scale",
  ylab = "log(Tau)",
  col = c("lightblue", "salmon"),
  outline = FALSE
)

stripchart(
  log(Tau) ~ Group, data = t1,
  vertical = TRUE, method = "jitter",
  add = TRUE, pch = 16,
  col = rgb(0, 0, 0, 0.5)
)

par(mfrow = c(1, 1))

# Each point is one participant.
# The line inside the box is the median.
# The box contains the middle 50% of observations; it is not a confidence interval.
# outline = FALSE avoids drawing outlier points twice.
# Those observations are still shown by stripchart() and retained in the analysis.
#
# INTERPRETATION:
# The original distributions are right-skewed.
# The log transformation compresses high values and improves symmetry.
# Disease tends to have higher values, but the groups overlap.


```

## 1.4 Residual Q-Q plots

Use model residuals to assess whether the log scale better supports the normal approximation.

```{r}
# 1.4 RESIDUAL Q-Q PLOTS

nfl_raw_model <- lm(NfL ~ Group, data = t1)
nfl_log_model <- lm(log(NfL) ~ Group, data = t1)

tau_raw_model <- lm(Tau ~ Group, data = t1)
tau_log_model <- lm(log(Tau) ~ Group, data = t1)

par(mfrow = c(2, 2))

qqnorm(residuals(nfl_raw_model), main = "NfL: raw residuals")
qqline(residuals(nfl_raw_model), col = "red")

qqnorm(residuals(nfl_log_model), main = "NfL: log residuals")
qqline(residuals(nfl_log_model), col = "blue")

qqnorm(residuals(tau_raw_model), main = "Tau: raw residuals")
qqline(residuals(tau_raw_model), col = "red")

qqnorm(residuals(tau_log_model), main = "Tau: log residuals")
qqline(residuals(tau_log_model), col = "blue")

par(mfrow = c(1, 1))

# Residual = observed value - fitted value.
# With Group as the only predictor, the fitted value is the group mean
# on the scale used in the model.
#
# The Q-Q plot compares ordered residuals with normal-distribution quantiles.
# A roughly straight pattern supports approximate normality.
# Log-scale residuals follow the reference line more closely here.
#
# A Q-Q plot does not prove normality or test the Healthy-Disease difference.
# It also does not check independence or equality of variances.
#
# These lm() models are used for diagnostics.
# Ordinary lm() coefficient tests assume a common residual variance.
# The Welch tests below allow unequal variances.


```

## 1.5 Welch test for NfL

Test NfL and express the mean-log difference as a Disease/Healthy geometric-mean ratio.

```{r}
# 1.5 WELCH TEST FOR NfL

nfl_disease <- log(t1$NfL[t1$Group == "Disease"])
nfl_healthy <- log(t1$NfL[t1$Group == "Healthy"])

nfl_test <- t.test(
  nfl_disease,
  nfl_healthy,
  alternative = "two.sided",
  var.equal = FALSE
)

nfl_test

nfl_ratio <- exp(nfl_test$estimate[1] - nfl_test$estimate[2])
nfl_ratio_CI <- exp(nfl_test$conf.int)

nfl_ratio
nfl_ratio_CI
nfl_test$p.value

# H0: equal population means of log(NfL).
# Equivalently, the Disease/Healthy geometric-mean ratio equals 1.
#
# Disease is the first argument, so the difference is Disease - Healthy.
# var.equal = FALSE selects Welch's test.
# It does not test whether the variances are equal.
#
# KEY OUTPUTS:
# t = 2.3473; df = 25.7209; p = 0.0268883.
# Difference in mean logs = 0.539618.
# Disease/Healthy geometric-mean ratio = 1.715351.
# Individual 95% CI for the ratio = [1.069115, 2.752211].
#
# REPORT SENTENCE:
# "Geometric-mean NfL was estimated to be 71.5% higher in Disease than Healthy
# participants (ratio 1.715, unadjusted 95% CI 1.069-2.752; Welch p = 0.027)."


```

## 1.6 Welch test for Tau

Repeat the same pre-defined comparison for Tau.

```{r}
# 1.6 WELCH TEST FOR TAU

tau_disease <- log(t1$Tau[t1$Group == "Disease"])
tau_healthy <- log(t1$Tau[t1$Group == "Healthy"])

tau_test <- t.test(
  tau_disease,
  tau_healthy,
  alternative = "two.sided",
  var.equal = FALSE
)

tau_ratio <- exp(tau_test$estimate[1] - tau_test$estimate[2])
tau_ratio_CI <- exp(tau_test$conf.int)

tau_test
tau_ratio
tau_ratio_CI
tau_test$p.value

# KEY OUTPUTS:
# t = 2.0689; df = 27.3590; p = 0.0481218.
# Disease/Healthy geometric-mean ratio = 1.719912.
# Individual 95% CI = [1.004820, 2.943909].
#
# REPORT SENTENCE:
# "Geometric-mean Tau was estimated to be 72.0% higher in Disease than Healthy
# participants (ratio 1.720, unadjusted 95% CI 1.005-2.944; Welch p = 0.048)."
#
# The interval is wide and its lower limit is very close to 1.
# Avoid rounding that lower limit to 1.00 when discussing significance.
#
# A p-value is not the probability that H0 is true.
# It describes how unusual the observed test result would be under H0
# and the assumptions of the test.


```

## 1.7 Consider the two biomarkers together

Account for analysing two biomarkers and state the exploratory conclusion.

```{r}
# 1.7 CONSIDER THE TWO BIOMARKERS TOGETHER

t1_p_values <- c(NfL = nfl_test$p.value, Tau = tau_test$p.value)

p.adjust(t1_p_values, method = "holm")

# OUTPUT: approximately 0.0537766 for both biomarkers.
#
# Both individual tests have p < 0.05.
# After Holm adjustment for the two tests, neither has adjusted p < 0.05.
# This does not prove equality or absence of an association.
# The individual confidence intervals above are not adjusted for multiplicity.
#
# FINAL TASK 1 STATEMENT:
# "Both biomarkers had higher estimated geometric means in Disease.
# However, both Holm-adjusted p-values were approximately 0.054.
# The analysis is exploratory, and the small Disease sample, wide confidence
# intervals and undocumented matching limit the conclusions.
# These comparisons do not establish diagnostic accuracy or causation."


# ============================================================================
```

**Key conclusion:** Both biomarkers had higher estimated geometric means in Disease. The Disease/Healthy ratios were 1.715 for NfL and 1.720 for Tau. Both individual p-values were below 0.05, but both Holm-adjusted p-values were approximately 0.054. The evidence is exploratory and limited by the small Disease sample and undocumented matching.

# Task 2 - Clinical study

**Research focus:** How many patients are needed to detect a 20% reduction in geometric-mean NfL?

```{r}
#| echo: false
#| output: false
# TASK 2 - PLAN AND ANALYSE A CLINICAL TRIAL
# ============================================================================

# QUESTION:
# How many patients are needed to detect a 20% reduction in geometric-mean
# NfL at three months?
#
# Both groups now have prion disease: Treatment versus Control.


```

## 2.1 Calculate the sample size

Use the Task 1 Disease-group variability to plan a two-arm trial.

```{r}
# 2.1 CALCULATE THE SAMPLE SIZE

sd_log <- sd(log(t1$NfL[t1$Group == "Disease"]))
sd_log

# OUTPUT: 0.745827.
# Use Disease participants because the new trial includes patients in both arms.
#
# A 20% reduction means GM_Treatment / GM_Control = 0.80.
# For example, reducing a geometric mean from 50 to 40:
# log(40) - log(50) = log(0.80) = approximately -0.22314.

sample_size <- power.t.test(
  delta = abs(log(0.80)),
  sd = sd_log,
  sig.level = 0.05,
  power = 0.80,
  type = "two.sample",
  alternative = "two.sided"
)

sample_size

n_per_arm <- ceiling(sample_size$n)
total_n <- 2 * n_per_arm

n_per_arm
total_n

# OUTPUT: approximately 176.33 per group.
# Round upwards to 177 per group: 354 participants in total.
#
# abs() uses the magnitude of the log difference.
# ceiling() rounds upwards to a whole number of participants.
#
# Alpha = 0.05 controls false positives under the null hypothesis.
# Power = 0.80 means about an 80% probability of detecting a difference when
# the true reduction is 20% and the planning assumptions are correct.
# It is not an 80% probability that the treatment works.
#
# power.t.test() assumes a common SD for independent groups.
# This is an approximate planning calculation for the later Welch analysis.
# The SD estimate comes from only 15 patients and is therefore uncertain.
# The required 354 participants must have usable outcomes; losses would
# require additional recruitment.
#
# REPORT SENTENCE:
# "Using the Disease-group SD of log(NfL), a two-sided significance level
# of 5% and 80% power, we recommend 177 evaluable participants per arm
# to detect a 20% reduction in geometric-mean NfL."


```

## 2.2 Specify the analysis plan before generating data

Fix the hypotheses, analysis scale and reporting strategy before generating data.

```{r}
# 2.2 SPECIFY THE ANALYSIS PLAN BEFORE GENERATING DATA

# Outcome: NfL at three months.
# Groups: Treatment and Control, assumed independent.
# Identified matched pairs are not supplied.
# Primary analysis: two-sided Welch t-test on log(NfL).
# Significance level: 0.05.
# H0: GM_Treatment / GM_Control = 1.
# H1: GM_Treatment / GM_Control differs from 1.
# Report: geometric-mean ratio, 95% CI and p-value.
#
# The target ratio of 0.80 is used for planning.
# The null hypothesis concerns a ratio of 1, meaning no difference.
#
# Check log-scale distributions and Q-Q plots within each arm.
# Report missing outcomes by arm and analyse available valid measurements.
# Acknowledge possible bias from missing outcomes.
# Investigate data errors, but do not remove valid observations merely
# because they look unusual.
#
# If assumptions fail seriously, report this and label any alternative
# analysis as a sensitivity analysis.
# Do not choose a test simply because it produces a smaller p-value.
#
# There is one primary outcome and one comparison, so no multiplicity
# correction is needed for the primary analysis.


```

## 2.3 Recruitment feasibility

Translate annual disease incidence into an approximate recruitment timeline.

```{r}
# 2.3 RECRUITMENT FEASIBILITY

# Population: 10 million.
# Incidence: 1-2 cases per million per year.
# This gives 10-20 new cases per year.
#
# If 20-80% meet the inclusion criteria:
# least favourable extreme = 10 * 0.20 = 2 eligible patients/year.
# most favourable extreme = 20 * 0.80 = 16 eligible patients/year.

eligible_per_year <- c(2, 16)
years_to_recruit <- total_n / eligible_per_year

years_to_recruit

# OUTPUT: 177 years at 2/year; 22.125 years at 16/year.
#
# This assumes that all eligible patients participate and have usable outcomes.
# Add three months of follow-up after the last participant is enrolled.
#
# REPORT SENTENCE:
# "Recruitment would take approximately 22-177 years at a single centre
# under the stated assumptions. The proposed single-centre study is unlikely
# to be feasible; recruitment across several regions would be needed."


```

## 2.4 Generate one trial

Generate one reproducible trial at the recommended sample size.

```{r}
# 2.4 GENERATE ONE TRIAL

library(CM2018rpackage)

# If hpd_trial_data() is unavailable, update the course package once:
# install.packages("devtools")
# devtools::install_github("adamdarwichkth/CM2018rpackage")
# Restart R after updating.

set.seed(20260910)
trial <- hpd_trial_data(n_per_arm)

head(trial)
table(trial$Group)
colSums(is.na(trial))

# OUTPUT: 177 in Group 0 (Control) and 177 in Group 1 (Treatment).
# The example contains no missing outcomes.
#
# The generator restarts ID numbering within each arm.
# These repeated labels do not identify matched patient pairs.
#
# set.seed() makes the example reproducible with the same generator version.

trial$Group <- factor(
  trial$Group,
  levels = c(0, 1),
  labels = c("Control", "Treatment")
)


```

## 2.5 Apply the pre-specified analysis

Apply the pre-specified log-scale Welch analysis to the generated trial.

```{r}
# 2.5 APPLY THE PRE-SPECIFIED ANALYSIS

treatment <- log(trial$NfL[trial$Group == "Treatment"])
control <- log(trial$NfL[trial$Group == "Control"])

trial_test <- t.test(
  treatment,
  control,
  alternative = "two.sided",
  var.equal = FALSE
)

trial_test

trial_ratio <- exp(trial_test$estimate[1] - trial_test$estimate[2])
trial_ratio_CI <- exp(trial_test$conf.int)

trial_ratio
trial_ratio_CI
trial_test$p.value

# OUTPUT for the previously checked generator and seed:
# Treatment/Control ratio = 0.749617.
# 95% CI = [0.678611, 0.828053].
# Welch p = 2.60014e-08.
#
# (1 - ratio) * 100 gives the estimated percentage reduction: about 25.0%.
# The CI corresponds to approximately 17.2% to 32.1% lower NfL.
#
# REPORT SENTENCE:
# "In the simulated trial, geometric-mean NfL was approximately 25% lower
# in Treatment than Control (ratio 0.750, 95% CI 0.679-0.828; p < 0.001)."
#
# These are simulated results, not observations from an actual clinical trial.

par(mfrow = c(1, 1))

boxplot(
  log(NfL) ~ Group, data = trial,
  outline = FALSE,
  col = c("lightblue", "salmon"),
  ylab = "log(NfL)",
  main = "Task 2: NfL at three months"
)

stripchart(
  log(NfL) ~ Group, data = trial,
  vertical = TRUE, method = "jitter",
  add = TRUE, pch = 16,
  col = rgb(0, 0, 0, 0.35)
)

par(mfrow = c(1, 2))

qqnorm(control, main = "Control: log(NfL)")
qqline(control, col = "blue")

qqnorm(treatment, main = "Treatment: log(NfL)")
qqline(treatment, col = "red")

par(mfrow = c(1, 1))

# Roughly straight Q-Q patterns support the normal approximation on the log scale.
# They do not test the treatment difference or establish independence.


```

## 2.6 Repeat the study

Repeat the trial to demonstrate the long-run meaning of power and sampling variability.

```{r}
# 2.6 REPEAT THE STUDY

set.seed(123)

p_values <- replicate(100, {

  new_trial <- hpd_trial_data(n_per_arm)

  result <- t.test(
    log(new_trial$NfL[new_trial$Group == 1]),
    log(new_trial$NfL[new_trial$Group == 0]),
    var.equal = FALSE
  )

  result$p.value
})

head(p_values)
sum(p_values < 0.05)
mean(p_values < 0.05)

# replicate() generates and analyses 100 separate trials.
# sum() counts significant results.
# mean() calculates the significant fraction because TRUE = 1 and FALSE = 0.
#
# With the generator checked for this exercise, the fraction is expected
# to be very close to 1. Use the output above to report your actual result.
# Even 100 significant results out of 100 would not prove exactly 100% power.
#
# WHY MIGHT THE FRACTION EXCEED 0.80?
# Planning used a ratio of 0.80 and SD_log = 0.745827.
# The checked course generator uses a ratio of 0.75 and SD_log = 0.4660144.
# A larger effect and smaller variability make detection easier.
#
# Generator source:
# https://github.com/adamdarwichkth/CM2018rpackage/blob/main/R/hello.R
#
# FINAL TASK 2 STATEMENT:
# "Task 1 was exploratory because the observed distributions informed our
# analysis choices. Task 2 follows a confirmatory approach within a simulation
# exercise: the outcome, sample size and analysis were specified beforehand.
# Replicated trials produce different estimates and p-values.
# Biomarker reduction alone does not establish improved clinical outcomes."


# ============================================================================
```

**Key conclusion:** Using the Disease-group SD of log(NfL), the planned study requires 177 evaluable participants per arm. At 2 to 16 eligible patients per year, single-centre recruitment would take approximately 22 to 177 years. The generated trial is simulated rather than real patient evidence.

# Task 3 - Dose-finding study

**Research focus:** Does each experimental dose change geometric-mean NfL compared with Control?

```{r}
#| echo: false
#| output: false
# TASK 3 - COMPARE THREE DOSES WITH CONTROL
# ============================================================================

# QUESTION:
# Does each dose change geometric-mean NfL compared with Control?
#
# Planned comparisons: Low-Control, Medium-Control and High-Control.
# We are interested in reductions, but use two-sided tests to allow for
# differences in either direction.


```

## 3.1 Import and describe the data

Check the four-arm dataset and compare descriptive summaries.

```{r}
# 3.1 IMPORT AND DESCRIBE THE DATA

# The first CSV column contains exported row numbers.
t3 <- read.csv("T3_data.csv", row.names = 1)

t3$Group <- factor(
  t3$Group,
  levels = c("Control", "Low", "Medium", "High")
)

str(t3)
table(t3$Group)
colSums(is.na(t3))
anyDuplicated(t3$ID)
any(t3$NfL <= 0)

# OUTPUT: 120 participants, 30 per group.
# No missing values, repeated IDs or non-positive NfL values.

aggregate(NfL ~ Group, data = t3, FUN = mean)
aggregate(NfL ~ Group, data = t3, FUN = median)

aggregate(
  NfL ~ Group,
  data = t3,
  FUN = function(x) exp(mean(log(x)))
)

# KEY OUTPUTS, in pg/mL:
# Group     Arithmetic mean    Median    Geometric mean
# Control        83.928        66.412         69.884
# Low            53.113        38.853         42.134
# Medium         46.084        43.262         39.361
# High           52.547        44.202         46.085
#
# All three dose groups have lower estimated geometric means than Control.
# These summaries do not yet account for sampling uncertainty.


```

## 3.2 Visualise original and log scales

Show individual observations on the original and log scales.

```{r}
# 3.2 VISUALISE ORIGINAL AND LOG SCALES

dose_colours <- c("grey80", "lightblue", "gold", "salmon")

par(mfrow = c(1, 2))

boxplot(
  NfL ~ Group, data = t3,
  outline = FALSE, col = dose_colours,
  main = "Task 3: original scale",
  ylab = "NfL (pg/mL)"
)

stripchart(
  NfL ~ Group, data = t3,
  vertical = TRUE, method = "jitter",
  add = TRUE, pch = 16,
  col = rgb(0, 0, 0, 0.4)
)

boxplot(
  log(NfL) ~ Group, data = t3,
  outline = FALSE, col = dose_colours,
  main = "Task 3: log scale",
  ylab = "log(NfL)"
)

stripchart(
  log(NfL) ~ Group, data = t3,
  vertical = TRUE, method = "jitter",
  add = TRUE, pch = 16,
  col = rgb(0, 0, 0, 0.4)
)

par(mfrow = c(1, 1))

# The original data are right-skewed.
# The log scale improves symmetry and supports proportional interpretation.


```

## 3.3 Global comparison: linear model and ANOVA

Use a global F-test to ask whether all four mean log(NfL) values could be equal.

```{r}
# 3.3 GLOBAL COMPARISON: LINEAR MODEL AND ANOVA

dose_model <- lm(log(NfL) ~ Group, data = t3)

anova(dose_model)

# Y = log(NfL).
# X = Group, a categorical predictor with four levels.
# Control is the reference group.
#
# OUTPUT: F(3, 116) = 5.3908; p = 0.001650.
#
# Global H0: all four population means of log(NfL) are equal.
# There is evidence that at least one mean differs.
# This does not identify which groups differ.
# It also does not imply that every pair differs.


```

## 3.4 Check model assumptions

Inspect residual normality and the variance pattern across the four groups.

```{r}
# 3.4 CHECK MODEL ASSUMPTIONS

par(mfrow = c(1, 2))

plot(
  fitted(dose_model),
  residuals(dose_model),
  pch = 16,
  xlab = "Fitted log(NfL)",
  ylab = "Residual",
  main = "Residuals versus fitted"
)

abline(h = 0, col = "red")

qqnorm(residuals(dose_model), main = "Task 3: residual Q-Q plot")
qqline(residuals(dose_model), col = "blue")

par(mfrow = c(1, 1))

tapply(log(t3$NfL), t3$Group, sd)

# OUTPUT:
# Control = 0.601; Low = 0.694; Medium = 0.599; High = 0.525.
#
# Ordinary ANOVA assumes common residual variance and approximately normal
# errors within groups. These SDs are reasonably similar; inspect the plots too.
#
# Four vertical bands in the residual plot are expected:
# the model gives one fitted mean to each of the four groups.


```

## 3.5 Planned comparisons with Holm correction

Compare each planned dose with Control while controlling multiplicity using Holm.

```{r}
# 3.5 PLANNED COMPARISONS WITH HOLM CORRECTION

# Reuse Welch's test from Tasks 1 and 2.
# Holm controls the false-positive risk across the three planned comparisons.
# These comparisons were defined before inspecting their p-values.

control_log <- log(t3$NfL[t3$Group == "Control"])
low_log <- log(t3$NfL[t3$Group == "Low"])
medium_log <- log(t3$NfL[t3$Group == "Medium"])
high_log <- log(t3$NfL[t3$Group == "High"])

low_test <- t.test(low_log, control_log, var.equal = FALSE)
medium_test <- t.test(medium_log, control_log, var.equal = FALSE)
high_test <- t.test(high_log, control_log, var.equal = FALSE)

# The dose is always the first argument.
# Exponentiating the difference therefore gives Dose/Control.

dose_results <- data.frame(
  Dose = c("Low", "Medium", "High"),

  Ratio = exp(c(
    low_test$estimate[1] - low_test$estimate[2],
    medium_test$estimate[1] - medium_test$estimate[2],
    high_test$estimate[1] - high_test$estimate[2]
  )),

  CI95_lower = exp(c(
    low_test$conf.int[1],
    medium_test$conf.int[1],
    high_test$conf.int[1]
  )),

  CI95_upper = exp(c(
    low_test$conf.int[2],
    medium_test$conf.int[2],
    high_test$conf.int[2]
  )),

  p_value = c(
    low_test$p.value,
    medium_test$p.value,
    high_test$p.value
  )
)

dose_results$p_Holm <- p.adjust(dose_results$p_value, method = "holm")

dose_results

# KEY OUTPUTS:
# Dose      Ratio      Individual 95% CI       Raw p       Holm p
# Low       0.6029     [0.4310, 0.8433]         0.003788     0.007575
# Medium    0.5632     [0.4130, 0.7681]         0.000475     0.001425
# High      0.6594     [0.4926, 0.8828]         0.005947     0.007575
#
# Use the Holm-adjusted p-values for decisions across the three comparisons.
# Holm remains valid when the comparisons share a control group.
# The displayed CIs are individual 95% intervals:
# p.adjust() adjusts p-values, not confidence intervals.
#
# ANOVA and these comparisons answer related questions using the same data.
# They are not independent replications.
#
# FINAL TASK 3 STATEMENT:
# "All three doses showed lower geometric-mean NfL than Control after Holm
# adjustment (all adjusted p < 0.01). Estimated reductions were 39.7% for Low,
# 43.7% for Medium and 34.1% for High.
# The pattern does not show a simple monotonic dose-response relationship.
# Although Medium had the lowest estimate, these comparisons do not establish
# that it is superior to the other active doses.
# Lower NfL alone does not establish clinical benefit."


# ============================================================================
```

**Key conclusion:** The global ANOVA found evidence that at least one mean log(NfL) differed. Low, Medium and High all had lower geometric-mean NfL than Control after Holm adjustment. Medium had the lowest point estimate, but the analysis did not establish superiority over the other doses.

# Task 4 - Clinical outcomes

**Research focus:** How does ordinal clinical response vary by treatment, and how is it related to NfL?

```{r}
#| echo: false
#| output: false
# TASK 4 - CLINICAL RESPONSE AND ITS RELATIONSHIP WITH NfL
# ============================================================================

# QUESTIONS:
# How do clinical response categories differ across treatment arms?
# How do those categories relate to NfL in the same participants?
#
# Response is ORDINAL.
# Categories have an order, but equal distances between categories are not defined.
# We do not calculate an arithmetic mean of the category labels.


```

## 4.1 Import and order the categories

Represent the clinical outcome as ordered categories and identify missing responses.

```{r}
# 4.1 IMPORT AND ORDER THE CATEGORIES

t4 <- read.csv("T4_data.csv", row.names = 1)

t4$Group <- factor(
  t4$Treatment,
  levels = c(0, 1, 2, 3),
  labels = c("Control", "Low", "Medium", "High")
)

response_levels <- c(
  "Asymptomatic",
  "No to slight symptoms",
  "Moderate disability",
  "Moderate to severe disability",
  "Severe disability"
)

t4$Response <- factor(
  t4$Response,
  levels = response_levels,
  ordered = TRUE
)

str(t4)
table(t4$Group)
table(t4$Response, useNA = "ifany")
colSums(is.na(t4))
anyDuplicated(t4$ID)

t4[is.na(t4$Response), c("ID", "Group")]

# OUTPUT: 120 participants, 30 per arm; no repeated IDs.
# Missing responses: ID 8 in Control and ID 98 in High.
#
# Observed category counts:
# Asymptomatic = 4.
# No/slight symptoms = 8.
# Moderate disability = 38.
# Moderate/severe disability = 25.
# Severe disability = 43.
# Missing = 2.
#
# Keep missing responses as NA.
# Clinical-response plots use the 118 observed responses.


```

## 4.2 Frequency and percentage tables

Calculate counts and percentages within each treatment arm.

```{r}
# 4.2 FREQUENCY AND PERCENTAGE TABLES

# Rows = response categories; columns = treatment groups.
response_counts <- table(t4$Response, t4$Group)

response_counts
colSums(response_counts)

# OUTPUT denominators: Control = 29; Low = 30; Medium = 30; High = 29.
# table() excludes missing responses by default.

response_percent <- 100 * prop.table(response_counts, margin = 2)

round(response_percent, 1)

# margin = 2 calculates percentages within columns, meaning within each group.
#
# COUNTS: Control / Low / Medium / High
# Asymptomatic                  1 /  1 /  1 /  1
# No to slight symptoms         1 /  1 /  3 /  3
# Moderate disability           8 / 13 / 10 /  7
# Moderate to severe            3 /  8 /  7 /  7
# Severe disability            16 /  7 /  9 / 11
#
# Severe disability percentages:
# Control = 55.2%; Low = 23.3%; Medium = 30.0%; High = 37.9%.


```

## 4.3 Visualise clinical response by treatment

Display the complete ordinal response distribution across the four arms.

```{r}
# 4.3 VISUALISE CLINICAL RESPONSE BY TREATMENT

response_colours <- c(
  "#1A9850", "#91CF60", "#FEE08B", "#FC8D59", "#D73027"
)

par(mfrow = c(1, 1), mar = c(5, 5, 4, 2))

barplot(
  response_percent,
  col = response_colours,
  ylim = c(0, 140),
  yaxt = "n",
  ylab = "Observed responses (%)",
  names.arg = c(
    "Control\n(n = 29)",
    "Low\n(n = 30)",
    "Medium\n(n = 30)",
    "High\n(n = 29)"
  ),
  main = "Task 4: clinical response by treatment"
)

axis(2, at = seq(0, 100, 20), las = 1)

legend(
  "top",
  legend = c(
    "Asymptomatic", "No/slight symptoms",
    "Moderate", "Moderate/severe", "Severe"
  ),
  fill = response_colours,
  ncol = 3,
  bty = "n",
  cex = 0.8
)

# Each bar represents 100% of available responses in that arm.
# The space above 100% is used only for the legend.
#
# Control has the largest proportion classified as Severe disability.
# The highest dose does not have the most favourable observed distribution.
# These are descriptive comparisons; no clinical-outcome hypothesis test
# has been performed.


```

## 4.4 Join Tasks 3 and 4 by participant ID

Join clinical response to the Task 3 biomarker by participant ID.

```{r}
# 4.4 JOIN TASKS 3 AND 4 BY PARTICIPANT ID

# Verify that each file contains unique IDs and the same set of participants.
stopifnot(
  anyDuplicated(t3$ID) == 0,
  anyDuplicated(t4$ID) == 0,
  setequal(t3$ID, t4$ID)
)

combined <- merge(
  t3,
  t4[, c("ID", "Treatment", "Response")],
  by = "ID"
)

nrow(combined)
# OUTPUT: 120.

# merge() links the same person across files even if row orders differ.
# Do not attach clinical responses by row position.

group_names <- c("Control", "Low", "Medium", "High")

all(
  as.character(combined$Group) ==
    group_names[combined$Treatment + 1]
)
# OUTPUT: TRUE; the treatment labels agree across the files.
# +1 converts treatment codes 0-3 into vector positions 1-4.

clinical <- combined[!is.na(combined$Response), ]

nrow(clinical)
# OUTPUT: 118 complete records for clinical-response plots.


```

## 4.5 Relate NfL to clinical severity

Visualise how NfL varies across increasing levels of clinical severity.

```{r}
# 4.5 RELATE NfL TO CLINICAL SEVERITY

aggregate(NfL ~ Response, data = clinical, FUN = median)

# Median NfL in pg/mL, from least to most severe category:
# Asymptomatic                 30.5955
# No to slight symptoms        34.1845
# Moderate disability          40.8800
# Moderate to severe           47.1310
# Severe disability            54.0970
#
# Overall medians increase with severity.
# Individual observations still overlap substantially.

short_response <- c(
  "Asymptomatic",
  "No/slight\nsymptoms",
  "Moderate",
  "Moderate/\nsevere",
  "Severe"
)

par(mfrow = c(1, 1), mar = c(6, 4, 4, 2))

boxplot(
  log(NfL) ~ Response,
  data = clinical,
  names = short_response,
  outline = FALSE,
  col = response_colours,
  cex.axis = 0.8,
  ylab = "log(NfL)",
  xlab = "",
  main = "Task 4: NfL across clinical response categories"
)

stripchart(
  log(NfL) ~ Response,
  data = clinical,
  vertical = TRUE,
  method = "jitter",
  add = TRUE,
  pch = 16,
  col = rgb(0, 0, 0, 0.45)
)

# Individual points show the small category sizes:
# Asymptomatic contains only 4 participants and No/slight symptoms contains 8.
# Boxes alone would hide this information.


```

## 4.6 Inspect the association within each treatment arm

Check whether the overall biomarker-response pattern is also visible within each treatment arm.

```{r}
# 4.6 INSPECT THE ASSOCIATION WITHIN EACH TREATMENT ARM

# Overall patterns can partly reflect differences between treatment groups.
# Repeat the same plot separately for each arm.

par(mfrow = c(2, 2), mar = c(6, 4, 3, 1))

for (group in levels(clinical$Group)) {

  group_data <- clinical[clinical$Group == group, ]

  boxplot(
    log(NfL) ~ Response,
    data = group_data,
    names = short_response,
    outline = FALSE,
    col = response_colours,
    cex.axis = 0.65,
    ylim = range(log(clinical$NfL)),
    ylab = "log(NfL)",
    xlab = "",
    main = paste(group, "- n =", nrow(group_data))
  )

  stripchart(
    log(NfL) ~ Response,
    data = group_data,
    vertical = TRUE,
    method = "jitter",
    add = TRUE,
    pch = 16,
    col = rgb(0, 0, 0, 0.45)
  )
}

par(mfrow = c(1, 1), mar = c(5.1, 4.1, 4.1, 2.1))

# The same y-axis range makes panels comparable.
# Within-arm patterns are variable and some categories contain very few patients.
# An overall association need not be equally strong within every treatment arm.
#
# Task 4 asks for visual exploration, so a formal ordinal model is not needed.
#
# FINAL TASK 4 STATEMENT:
# "Severe disability was more frequent in Control (55.2%) than in the
# treatment arms (23.3-37.9%) among observed responses.
# Across all arms, median NfL increased with clinical severity, although
# biomarker values overlapped and within-arm patterns were variable.
# These findings are descriptive and exploratory.
# Two missing responses and small category sizes limit interpretation.
# The visual associations do not establish causation, treatment efficacy
# on clinical outcomes, or NfL as a validated surrogate."


# If you uncommented pdf() at the top, also uncomment the following line:
# dev.off()
```

**Key conclusion:** Severe disability was observed most frequently in Control. Across all arms, median NfL increased with clinical severity, but observations overlapped and within-arm patterns were variable. The visual association does not establish causation or clinical efficacy.
